---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 8,
  dpi = 300,
  out.width = "75%"
)
```

# flow field art(?)

## Introduction
* A few worked examples of a method for creating generative art from flow fields in R
* This is a **proof of concept and is not rigorously tested or optimised**
* It's just for fun and hopefully it can be a useful jumping off point for other people to develop further (and please let me know if you do!)

Add tidyverse packages to search path
```{r warning = FALSE, message = FALSE}
library(tidyverse)
```

## Noise generation
* I'll start by generating the noise field using the `{ambient}` package
* I also define some canvas dimensions `x_side` and `y_side` and set a seed for reproducibility
* The noise field matrix is re-scaled to range between -90 and 90
  * These are the angles that define the direction that flow lines will travel at that point in the flow field
* Also create a dataframe of the flow field matrix in order to visualise it with ggplot2
  * This is confusing for me but the dataframe version will have a reversed y-axis because a y value of 1 is the bottom row of a plot but it is the top (or first index) row of the matrix. So I will need to plot with a reversed y-axis to get it the right way around (possibly not too important if just making artistic plots of noise)
  
```{r}
global_seed <- 4
set.seed(global_seed)
x_side <- 400
y_side <- 400

# Create a noise field matrix
m <- 
  ambient::noise_simplex(
    c(y_side, x_side),
    frequency = 0.004,
    octaves = 2,
    pertubation = "normal",
    pertubation_amplitude = 2,
    fractal = 'billow') |> 
  scales::rescale(c(-90, 90)) # scale noise values to angles in degrees (-90 to 90)

# Create a dataframe of the noise matrix for demo visualisation
# (remembering that the y axis will be reversed)
m_df <- crossing(x=1:ncol(m),  y=1:nrow(m)) |> mutate(angle = as.vector(m))
```

* Visualise the noise flow field
```{r}
ggplot() +
  geom_raster(data = m_df, aes(x, y, fill = angle))+
  coord_equal()+
  scale_fill_gradient(low = "black", high = "white")+
  scale_y_reverse()
```

## Flow line generation
* Now I'll define a function that takes a flow field `angle_matrix` (like `m` which I made above)  and a bunch of other parameters, and returns the x and y coordinates of the polygon that makes a tapered flow field line
```{r}
#' Get the coords of a tapered flow line polygon
#' 
#' The x_start and y_start will be used to index columns and rows of the angle_matrix
#' This means y values from a conventional Cartesian coordinate system will be reversed as y = 1 will be row 1 of the matrix 
#'
#' @param x_start x starting point of flow line on angle_matrix
#' @param y_start y starting point of flow line on angle_matrix
#' @param step_length step length
#' @param n_steps number of steps
#' @param angle_matrix matrix of angles (the field)
#' @param taper_min min value of taper size
#' @param taper_max max value of taper size
ff_polys <- function(
    x_start, 
    y_start, 
    step_length, 
    n_steps, 
    angle_matrix, 
    taper_max, 
    taper_min){
  
  # Initialise vectors with the starting x and y values filled with NAs
  out_x <- c(x_start, rep(NA, n_steps))
  out_y <- c(y_start, rep(NA, n_steps))
  
  # If the starting point is outside the angle_matrix dimensions, return NULL
  if(x_start > ncol(angle_matrix) |
     x_start < 1 |
     y_start > nrow(angle_matrix) |
     y_start < 1){
    return(NULL)
  }
  
  # Loop through each step as we travel across the angle matrix
  for(i in 1:n_steps){
  
    # Get the angle of the nearest flow field point where we are for this iteration
    a <- angle_matrix[round(out_y[i]), round(out_x[i])]
    
    # Compute how far to move in x and y for the given angle and step_length
    step_x <- cos(a*(pi/180))*step_length
    step_y <- sin(a*(pi/180))*step_length
    
    # Add the distance in x and y to the current location
    next_x <- out_x[i] + step_x
    next_y <- out_y[i] + step_y
    
    # If the next point in the path sits outside the angle matrix, stop iterating along the path
    if(next_x > ncol(angle_matrix) |
       next_x < 1 |
       next_y > nrow(angle_matrix) |
       next_y < 1){
      break
    }
    
    # Append the new x and y location to the output 
    # (ready to be used as the starting point for the next step iteration)
    out_x[i+1] <- next_x
    out_y[i+1] <- next_y
  }
  
  # Return tibble of the x, y, paths
  # The polygon goes out along x and then back along rev(x)
  # The y values have a taper added to them on the way out (along x)
  # and then have a reverse taper subtracted from them on the way back (along rev(x))
  tibble(x = c(out_x, rev(out_x)),
         y = c(out_y + seq(taper_min, taper_max, l=length(out_y)), 
               rev(out_y) - seq(taper_max, taper_min, l=length(out_y)))) |> 
    # Finally remove any NA entries from the path where the flow line got to the
    # edge of the angle_matrix and we stopped iterating
    filter(!is.na(x), !is.na(y))
}
```

## Demo example
* A small example to demonstrate lines on the flow field
* Define the starting points of 15 flow lines and their various parameters
  * With `step_length = 1` and `n_steps = 200` the flow lines should be 200 units long
  * Tapering from 0 to 5 units thick

* Run the `ff_polys` function for each flow line starting point, passing the same angle matrix `m`
* Then unnest each flow line into a long format dataframe containing the x and y coordinates of each polygon
```{r}
set.seed(global_seed)

# Define starting points and flow line parameters
starting_params <-
  tibble(
    x_start = runif(15, 1, ncol(m)),
    y_start = runif(15, 1, nrow(m)))  |> 
  mutate(
    id = row_number(),
    step_length = 1,
    n_steps = 200,
    taper_max = 5,
    taper_min = 0)

# Compute the flow line polygon coordinates
flow_field_coords <-
  starting_params |> 
  mutate(
    paths = pmap(
      .l = list(x_start = x_start,
                y_start = y_start,
                step_length = step_length,
                n_steps = n_steps,
                taper_max = taper_max,
                taper_min = taper_min),
      .f = ff_polys,
      angle_matrix = m)) |> 
  unnest(cols=paths)
```

* Visualise the flow lines over the angle matrix
* Note that as the tapering is applied as a subtraction and addition to the y-coordinate of the flow line, if a line goes back on itself in x, the tapering will create a polygon that crosses itself
* We see that the flow lines are approximately 200 units long
```{r}
ggplot() +
  geom_raster(data = m_df, aes(x, y, fill = angle))+
  geom_polygon(
    data = flow_field_coords, 
    aes(x, y, group = id), 
    col = 1, 
    fill = "green",
    alpha = 1/2)+
  geom_point(data = starting_params, aes(x_start, y_start), size = 2, col = "black")+
  coord_equal()+
  scale_fill_gradient2()+
  scale_y_reverse()
```

## Art output
* Now run again with more flow lines and some personal choices for aesthetic qualities
* I have chosen to fill each polygon based on its start point relative to the centre of the angle matrix
```{r}
set.seed(global_seed)
n_flow_lines <- 2000

# Define starting points and flow line parameters
starting_params <-
  tibble(
    x_start = runif(n_flow_lines, 1, ncol(m)),
    y_start = runif(n_flow_lines, 1, nrow(m))) |> 
  mutate(
    id = row_number(),
    step_length = 0.1,
    n_steps = 5000,
    taper_max = 5,
    taper_min = 0)

# Compute flow line polygon coordinates
flow_field_coords <-
  starting_params |> 
  mutate(
    paths = pmap(
      .l = list(x_start = x_start,
                y_start = y_start,
                step_length = step_length,
                n_steps = n_steps,
                taper_max = taper_max,
                taper_min = taper_min),
      .f = ff_polys,
      angle_matrix = m)) |> 
  unnest(cols=paths)

# Render plot
ggplot() + 
  geom_polygon(
    data = flow_field_coords, 
    aes(x, y, group = id, 
        fill = sqrt((nrow(m)/2 - y_start)^2 + (ncol(m)/2 - x_start)^2)),
    col = NA,
    alpha = 0.4)+
  coord_equal()+
  scale_fill_viridis_c(option = "plasma", direction = -1)+
  scale_y_reverse()+
  theme_void()+
  theme(
    panel.background = element_rect(fill = "black", color = NA),
    legend.position = "")
```


## Using an image as a flow field

### Read image
```{r out.width = "20%"}
library(magick)
img <- image_read('https://i.pinimg.com/564x/e2/43/c0/e243c0f7b9ee95151d2f2c045367047c.jpg')
img
```

### Process image
* Define some parameters for imahe size and the number of flow lines wanted
* Resize image, greyscale, blur, flip and convert to matrix rescaled to angle range (0 to 180 degrees in this case)
```{r}
x_side <- 200
n_flow_lines <- 4000

m <-
  img |> 
  image_resize(paste0(x_side, "x")) |>
  image_convert(colorspace = "gray") |>
  image_flip() |>
  image_despeckle(2) |>
  image_blur(radius = 4, sigma=5) |>
  image_raster(tidy = FALSE) |>
  col2rgb() |>
  magrittr::extract(1,) |>
  matrix(ncol = x_side, byrow = TRUE) |>
  scales::rescale(c(0, 180))
```

* Generate flow lines and plot as above
```{r}
set.seed(global_seed)
starting_params <-
  tibble(x_start = runif(n_flow_lines, 1, ncol(m)),
         y_start = runif(n_flow_lines, 1, nrow(m))) |>
  mutate(id = row_number(),
         step_length = 1,
         n_steps = 300,
         taper_max = 5,
         taper_min = 0)

flow_field_coords <-
  starting_params |> 
  mutate(paths = pmap(
    .l = list(x_start = x_start,
              y_start = y_start,
              step_length = step_length,
              n_steps = n_steps,
              taper_max = taper_max,
              taper_min = taper_min),
    .f = ff_polys,
    angle_matrix = m)) |> 
  unnest(cols=paths)

ggplot() + 
  geom_polygon(
    data = flow_field_coords, 
    aes(x, y, group = id, 
        fill = sqrt((nrow(m)/2 - y_start)^2 + (ncol(m)/2 - x_start)^2)),
    col = NA,
    alpha = 0.4)+
  coord_equal()+
  scale_fill_viridis_c(option = "mako", direction = -1)+
  theme_void()+
  theme(
    panel.background = element_rect(fill = "black", color = NA),
    legend.position = "")
```
